---
title: "Homework 3"
author: "Graham Miotke"
date: "10/4/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
library(lubridate)
```

Link to github: https://github.com/gmiotke/433-Homework-3

Introduction:

The question we were tasked with finding is what time of the day should you fly if you are looking to minimize delays. There are many different factors which go into this question. One could look for the proportion of canceled flights, how long the flights were delayed, or a combination of the two. I felt like it would be most beneficial to look at the average delay. Without taking into account any other factors, we can see that the best time to fly is early in the morning, with very minimal delays. After the early morning hours, delay times steadily climb until the evening where they are at a peak. Eventually, delay times do seem to come down, but only once you get to 10 pm and later.

```{r}
flights%>%
  mutate(dep_delay = (dep_delay%/%100) * 60 + (dep_delay%%100))%>%
  group_by(hour)%>%
  summarize(avg_delay = mean(dep_delay, na.rm = T),
            num_flights = n())%>%
  arrange(avg_delay)%>%
  ggplot()+geom_line(aes(hour,avg_delay))+geom_point(aes(hour,avg_delay))

```

After determining this, I decided to check whether delay times were related to the season. In general, each season followed a similar pattern to the non-impacted results, with some seasons seeing worse delay times than others. Next, I wanted to see if the delay times were in any way related to the airlines in which people decided to fly. These results were a bit more scattered, but most still seemed to have an increase as the day went on. These increases seemed to be much more marginal for some, but it seemed like the morning was still the best time to fly and the evening was the worst for most airlines. Finally, I wanted to see if different weather variables impacted the optimal flying time. Based on the wind speed and temperature variables, it would seem that the greatest delays actually come a bit earlier in the day, but the best time to fly remain the early morning hours.

Best time to fly based on season:

```{r}
flights%>%
  mutate(dep_delay = (dep_delay%/%100) * 60 + (dep_delay%%100),
         season = case_when(
           month == 12|month == 1|month == 2 ~ "Winter",
           month == 3|month == 4|month == 5 ~ "Spring",
           month == 6|month == 7|month == 8 ~ "Summer",
           month == 9|month == 10|month == 11 ~ "Fall"
         ))%>%
  group_by(season, hour)%>%
  summarize(avg_delay = mean(dep_delay, na.rm = T),
            num_flights = n())%>%
  ggplot()+geom_line(aes(hour,avg_delay))+geom_point(aes(hour,avg_delay))+facet_wrap(~season)


```

While the seasons did not drastically change the conclusion that the morning hours are the best time to fly, there are some staggering differences between seasons. Fall has a very consistent average delay time, meaning there seems not to be a bad time to schedule a flight. The next best would be winter. While the average delay time in winter does increase, it is not nearly as high as in the spring or summer. The winter months do have quite low delays in the early hours compared to the other seasons. Spring has a continuous incline, but a sharp decline after the 21st hour. The delays in the spring are worse than the fall and winter for almost all hours of the day. Summer has the greatest change from the early hours to the later hours. In summer, once you reach about 3 pm the delays are extremely high, meaning if you want to avoid long delays, it is necessary to fly in the morning. All in all, the seasons all follow a similar pattern, the difference between maximum and minimum delay times, however, does vary greatly between seasons

Best time to fly based on carrier:

```{r}
fa = flights%>%
  mutate(dep_delay = (dep_delay%/%100) * 60 + (dep_delay%%100))%>%
  group_by(carrier, hour)%>%
  summarize(avg_delay = mean(dep_delay, na.rm = T),
            num_flights = n())%>%
  arrange(avg_delay)%>%
  filter(num_flights > 10)%>%
  left_join(airlines, by = "carrier")

head(fa)
tail(fa)

ggplot(fa)+geom_line(aes(hour,avg_delay))+geom_point(aes(hour,avg_delay))+facet_wrap(~carrier)

```

It is a bit harder to analyze whether the best time to fly relates to the airline which you choose to fly. This is partially because many airlines have too few flights during a certain time period to make the data reliable. This leaves some airlines with very few data points, such as Frontier Airlines (F9). Of the airlines with consistent data points, there were multiple different trends. Some airlines, such as Delta (DL) and US have very consistent delays throughout all hours of the day. There are other airlines which follow the general trend of low delays in the morning and higher delays in the evening, such as Southwest (WN) and ExpressJet (EV). So, depending on the airline you choose ot fly with, you could see consistent delay times, making no one time the best to fly. You could also potentially see increasing delay times throughout the day meaning you should fly in the morning.

Best Time to Fly Based on Weather:

```{r}
weather2 = weather%>%
  mutate(date = make_date(year,month,day))

fw = flights%>%
  mutate(date = make_date(year,month,day))%>%
  left_join(weather2)%>%
  group_by(hour,origin)%>%
  summarize(avg_delay = mean(dep_delay, na.rm = T),
            mean_temp = mean(temp, na.rm = T),
            mean_ws = mean(wind_speed, na.rm = T),
            num_flights = n())
  
ggplot(fw, aes(hour,mean_ws, color = origin))+geom_point()
ggplot(fw, aes(mean_ws,avg_delay))+geom_point()+geom_smooth(se= F, method = "lm")+facet_wrap(~origin)
ggplot(fw, aes(hour,mean_temp, color = origin))+geom_point()
ggplot(fw, aes(mean_temp,avg_delay))+geom_point()+ geom_smooth(se=F, method= "lm")
```

The final relationship I attempted to find was the relationship between best time to fly and weather. I decided to use the variables temperature and wind speed because it seemed like these factors could impact whether it was safe for a plane to take off. First I looked at wind speed. Surprisingly, on average it seemed like the wind speed was slowest in the early morning hours and late night hours, but fastest in the afternoon and early evening. After that, it was found that there is a positive correlation between wind speed and average delay. Therefore, it would appear based on the wind speed that the best time to fly would be when the wind speed is lowest, so in the early morning and later night hours. I also found that the different origins handle the wind speed differently. While they all seem to handle lower wind speeds with similar delays, JFK actually seems to handle the highest wind speeds with the lowest delay of the three origin airports. Next, I wanted to see if temperature had any impact on the average delay time. I found that in general the trend in temperature was very similar to that of wind speed. Surprisingly, there was also a positive correlation between mean temperature and average delay. This could be due to many factors, such as number of flights. However, this finding also confirms the idea that the early morning hours and the later night hours are the optimal time to schedule a flight. Unlike wind speed, temperature did not seem to differ much between the different origin airports. In conclusion, weather would seem to suggest that the best flying time is still the morning and late night, but the worst time to schedule a flight is actually a bit earlier than the other analysis suggests.


